Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex
Final pass through SC2

One of the unusual properties of this dataset is that whole genome alignments
are stored, such that information on all 29,903 bases is stored for every
sample. Given the number of samples and that nearly all bases have 
variation (all sites have at least one sample with missing data,
and 99.4\% have more than one allele)
it is simply more convenient to
store the entire alignment for each sample than to manage the complexity
of subsetting the sites. Similarly, the usual convention in VCF Zarr
is that the \texttt{variant\_alleles} field contains the REF and ALT
alleles as observed (such that the zero'th allele is always the reference).
In this case it is more convenient to encode the alleles as the original
bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
such that the allele 0 is always A, 1 is C etc.
Other than these minor deviations in interpretation, the Zarr follows
the VCF Zarr specification and can be processed using the same tools.

\begin{table}
\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
whole genome alignments and metadata 
(4,384,310 samples, 29903 variants)
consisting of 37 fields and 391 MiB of storage, 
of which 280MiB is used by aligments ($\sim 81$X smaller 
than gzipped FASTA). The chunk size is 100 variants $\times$
10,000 samples.
Shown are the top 9 fields consuming at least 1\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
Field names (derived from source CSV file) are shortened for 
display purposes by replacing ``Genbank'' with ``G''
and ``Viridian'' with ``V''.
\label{tab-sc2-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
\bottomrule
\end{tabular}
\end{table}

Decompressed, the Viridian dataset consists of 48
FASTA files totaling 125 GiB of storage and a 1.4 GiB
TSV file for sample metadata.
When aligned to the reference and converted to Zarr, the
entire dataset is stored in a single 391 MiB Zip file.
When compressed with bgzip (so they can be indexed; see next 
paragraph) the alignments require 22 GiB of storage.
Table~\ref{tab-sc2-data} summarises the storage used by individual
arrays, and shows that the alignment data takes up 298 MiB. 
This is around 81X smaller than gzipped FASTA.
The original dataset is distributed as xz compressed FASTA files, which
achieves even higher compression levels at a total of 218 MiB.
However, these files cannot be accessed in-situ and must be decompressed
in order to access the data.

Computational accessibility for the Zarr is excellent. 
Retrieving the full alignment for the 1 millionth sample required
[x ms]. Performing the same operation on the corresponding indexed FASTA
file using pyfaidx~\citep{shirley2015efficient} required [x ms].
Similarly, accessing calls for all samples 
at the 10,000th site requires xms
The storage of metadata and genetic data in a single file also
makes many queries efficient and convenient. For example, computing the
average number of deletions in samples with a Scorpio designation of Alpha
%TODO "x" on next line
required x seconds and 5 lines of code using Dask.

>>>>

One of the unusual properties of this dataset is that whole genome alignments
are stored, such that information on all 29,903 bases is stored for every
sample. Given the number of samples and that nearly all bases have 
variation (all sites have at least one sample with missing data,
and 99.4\% have more than one allele)
it is simply more convenient to
store the entire alignment for each sample than to manage the complexity
of subsetting the sites. Similarly, the usual convention in VCF Zarr
is that the \texttt{variant\_alleles} field contains the REF and ALT
alleles as observed (such that the zero'th allele is always the reference).
In this case it is more convenient to encode the alleles as the original
bases and IUPAC uncertainty codes~\citep{cornish1985nomenclature}, 
such that allele 0 is always A, 1 is C etc.
Other than these minor deviations in interpretation, the Zarr follows
the VCF Zarr specification and can be processed using the same tools.

\begin{table}
\caption{Summary for VCF Zarr conversion of Viridian SARS-CoV-2
whole genome alignments and metadata 
(4,384,310 samples, 29903 variants)
consisting of 37 fields and 391 MiB of storage, 
of which 280MiB is used by aligments ($\sim 81$X smaller 
than gzipped FASTA). The chunk size is 100 variants $\times$
10,000 samples.
Shown are the top 9 fields consuming at least 1\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
Field names (derived from source TSV file) are shortened for 
display purposes by replacing ``Genbank'' with ``G''
and ``Viridian'' with ``V''.
\label{tab-sc2-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_genotype & int8 & 279.38 MiB & 447.5 & 80.04\% \\
sample\_G\_tree\_name & str & 13.18 MiB & 2.5 & 3.78\% \\
sample\_G\_accession & str & 7.31 MiB & 4.6 & 2.09\% \\
sample\_Sample & str & 7.26 MiB & 4.6 & 2.08\% \\
sample\_Experiment & str & 6.9 MiB & 4.8 & 1.98\% \\
sample\_id & str & 6.83 MiB & 4.9 & 1.96\% \\
sample\_V\_N & int16 & 4.32 MiB & 1.9 & 1.24\% \\
sample\_V\_pangolin\_1.29 & str & 3.98 MiB & 8.4 & 1.14\% \\
sample\_V\_pangolin & str & 3.89 MiB & 8.6 & 1.11\% \\
\bottomrule
\end{tabular}
\end{table}

Decompressed, the Viridian dataset consists of 48
FASTA files totaling 125 GiB of storage and a 1.4 GiB
TSV file for sample metadata.
When aligned to the reference and converted to Zarr, the
entire dataset is stored in a single 391 MiB Zip file.
When compressed with bgzip (so they can be indexed; see next 
paragraph) the alignments require 22 GiB of storage.
Table~\ref{tab-sc2-data} summarises the storage used by individual
arrays, and shows that the alignment data requires 298 MiB,
around 81X smaller than gzipped FASTA.
The original dataset is distributed as xz compressed FASTA files, which
achieves even higher compression levels at a total of 218 MiB.
However, these files must be decompressed in order to access the data
using standard FASTA processing techniques.

Computational accessibility for the Zarr is excellent. For example,
computing the total number of missing or ambiguous bases for 
all 4.4 million samples required 55.5 seconds elapsed (6min 45s CPU)
using 8 threads. This required 11 lines of code, using only the 
Python standard library and NumPy. Similarly, computing allele
counts at each site required 4min 53s (25min 35s CPU) using a slightly
more complicated 16 line Python function.
Existing methods for working with FASTA data provided efficient 
access to individual alignments using the fai index format, which
works with uncompressed or bgzip compressed files.
For example, retrieving the alignment for the 1 millionth sample 
from the corresponding chunk file requires only a few milliseconds
using pyfaidx~\citep{shirley2015efficient} or 
pysam~\citep{pysam}. Retrieving the same alignment using Zarr 
required 365ms. 
Accessing calls for all samples at at a given 
site requires a similar time (here, accessing site 14,951
required 276ms). 
FASTA cannot support efficient retrieval
of data for a particular site in a set of alignments.
Storing data chunked in two dimensions means that we can avoid
the classic choice between sample-major or variant-major storage,
and access it efficiently in both dimensions.
